Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2014-11-18 manuscript.tex
Cite IPython. Connect algorithm with code
Mention the code parts in the algorithm description. Mention notebooks
for reproducing results and cite IPython.

@@

\begin{equation}
    \frac{d}{L_i} < D,
    \label{eq:condition}
\end{equation}
\noindent
where $i \in (\lambda, \phi, r)$
and $D$ is a positive scalar called the ``distance-size ratio''.
$D$ determines how close the
computation point can be before we must divide the tesseroid.

We used a stack-based implementation of the algorithm instead of using
recursive function calls.

@@

The algorithm is described for a single tesseroid and computation point.

The recursive division algorithm implemented with a stack is as follows.

Start by initializing the stack with the tesseroid.

Thus, so far, the stack has only one element.

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last element from it).
    \item Compute the distance $d$ between the tesseroid and
        the computation point (equation~\ref{eq:distance}).
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ (equations~\ref{eq:sizelon}-\ref{eq:sizer}).
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature with order two
        (equation~\ref{eq:glq3d}).
    \item If any of the dimensions fail the condition,
        divide the tesseroid in half along each dimension that failed the
        condition.
        Then, place the smaller tesseroids into the stack.
    \item Repeat the above steps until the stack is empty.
\end{enumerate}

>>>>

All figures and numerical experiments in this article were produced by the
IPython notebook\footnote{http://ipython.org}
files \citep{Perez2007} in the supplementary material.

These files combine the analysis source code
with explanatory text, equations, and the figures generated by the code.

This supplementary material can be run and executed locally or viewed
online\footnote{https://github.com/pinga-lab/paper-tesseroids}
(in a non-executable form).

@@

\begin{equation}
    \frac{d}{L_i} < D,
    \label{eq:condition}
\end{equation}
\noindent
where $i \in (\lambda, \phi, r)$
and $D$ is a positive scalar called the ``distance-size ratio''.

$D$ determines how close the
computation point can be before we must divide the tesseroid.

Thus, the value of $D$ is indirectly responsible for
the accuracy of the solution and the computation time.

The modeling programs provide default values of $D$.

It can also be controlled through command-line arguments passed to the
programs.

We used a stack-based implementation of the algorithm instead of using
recursive function calls, as  originally proposed by \citet{Li2011}.

@@

We will describe the stack-based algorithm for a single tesseroid and
computation point.

The algorithm is implemented in function \texttt{calc\_tess\_model\_adapt}
of the file \texttt{src/lib/grav\_tess.c}.

Start by creating an empty stack of tesseroids.

The  stack is represented by the \texttt{stack} variable,
an array of \texttt{TESSEROID} structures.

We must define a maximum size for the stack.

We use the integer \texttt{stktop} to keep track of the index of the last
element in the stack (the top of the stack).

The stack is initialized with the input tesseroid.

The initialization is done by copying the tesseroid into the stack and
setting \texttt{stktop} to zero (the first element).

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last element from it).
        This will cause \texttt{stktop} to be reduced by one.
    \item Compute the distance $d$ between the tesseroid and
        the computation point (equation~\ref{eq:distance}).
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ (equations~\ref{eq:sizelon}-\ref{eq:sizer}).
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature with order two
        (equation~\ref{eq:glq3d}).
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number new (smaller) tesseroids fit in the stack.
             In other words, the number of new elements plus \texttt{stktop}
             must be smaller than the maximum stack size.
             If the stack is full, warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\texttt{stktop = -1}).
\end{enumerate}

This stack-based implementation has some advantages over the original recursive
implementation suggested by \citet{Li2011}.

(1) It will be faster because it bypasses the overhead of function calls.

The speedup will be specially significant if the implementation is done
in high level interpreted languages, like Matlab and Python.

(2) It gives the developer more control over the recursion step.

In recursive implementations,
the developer has no control over the maximum number of consecutive
recursive calls (the ``recursion depth'').

This limit may vary with programming language, compiler, and operating system.

Overflowing the maximum recursion depth usually results in program
crashes with cryptic or inexistent error messages.

In the stack-based implementation, the developer has complete control.

Overflowing of the stack can be handled gracefully with an error message or
even performing a suitable approximation of the result.
